Verifying a Postfix Language \\ \large CS595 Final Project Proposal

### 2021-11-17 project/proposal.tex
Update proposal.tex

\documentclass{article}
\usepackage[english]{babel}

% Set page size and margins
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

@@

@@

    \item \textbf{Lambdas:} Lambdas aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.

@@

    # Define a recursive factorial lambda
    ((I32)(I32) rec:
        $n =
        (: 1 ) (: 1 ) $branch fun
        (: n 1 > ) (: n 1 - fac n * ) $branch fun
        branch
    ) $fac =

@@

\subsection{Lambdas}
Via the same syntax already defined

@@

\subsection{Operators, Keywords and Constants}
So in summary our builtin operators would be: 
\begin{itemize}
    \item \texttt{+} - numeric add
    \item \texttt{*} - numeric mul
    \item \texttt{if-then-else} - ternary operator
    \item \texttt{unpack} - destruct tuple (an attempt will be made)
    \item \texttt{\~} - Extract value from an escaped identifier without invoking it (less important for safety but useful for defining additional operators using these as base)
    \item \texttt{=} - the expression \texttt{e \$x = ...} is equivalent to \texttt{let x = e in ...}. But note that there are no 'bound variables' in this language.
    \item \texttt{type} - get type of argument (equivalent to \texttt{type()} in python
    \item \texttt{==} - current plan is to have an ambiguous == operator which accepts either 2 types or 2 equivalently typed values like in the language
    \item \texttt{|} - used to form type unions
    \item \texttt{Arrow} - lambda types
    \item \texttt{Unit} - keyword to distinguish from the empty tuple ()
    \item \texttt{Void} - used due to lack of syntax for empty union
\end{itemize}

@@

\subsubsection{Classes}
It would be nice to include classes in our language (notably recursive classes). However this would involve subtyping and some other concepts so might not have time to add them to the model

As much as we would like to include classes (notably recursive classes), it would likely add excess complexity, requiring subtyping (on N dimensions); untyped lambdas which can operate on types; and likely some other painpoints. And it's not clear how they could make the system unsafe as apart from being used to define recursive types are simply specialized versions of preexisting types.

\subsubsection{Functions}
Really the only reason I opted for the function overloading system instead of normal branching in the original language was in order to make it easier to extend the language.
Function overloading seems like it would be needlessly painful to formalize and doesn't contribute to our proof of soundness any more than branching via a ternary.

@@

\subsection{Findings, Resolution?, Conclusions}
It's entirely possible that we could encounter problems with the language design, although pretty unlikely since we're even requiring type signatures on lambdas.

>>>>

\documentclass{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with`a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

@@

\input{macros}

@@

    \item \textbf{closures:} closures aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.

@@

    # Define a recursive factorial closure
    ((I32)(I32) rec:
        $n =
        (: 1 ) (: 1 ) $branch fun
        (: n 1 > ) (: n 1 - fac n * ) $branch fun
        branch
    ) $fac =

@@

\subsection{closures}
Via the same syntax already defined

@@

\subsection{Operators, Keywords and Constants}
So in summary our builtin operators would be: 
\begin{itemize}
    \item \texttt{+} - numeric add
    \item \texttt{*} - numeric mul
    \item \texttt{if-then-else} - ternary operator
    \item \texttt{unpack} - destruct tuple (an attempt will be made)
    \item \texttt{\~} - Extract value from an escaped identifier without invoking it (less important for safety but useful for defining additional operators using these as base)
    \item \texttt{=} - the expression \texttt{e \$x = ...} is equivalent to \texttt{let x = e in ...}. But note that there are no 'bound variables' in this language.
    \item \texttt{type} - get type of argument (equivalent to \texttt{type()} in python
    \item \texttt{==} - current plan is to have an ambiguous == operator which accepts either 2 types or 2 equivalently typed values like in the language
    \item \texttt{|} - used to form type unions
    \item \texttt{Arrow} - closure types
    \item \texttt{Unit} - keyword to distinguish from the empty tuple ()
    \item \texttt{Void} - used due to lack of syntax for empty union
\end{itemize}

@@

\subsubsection{Classes}
As much as we would like to include classes (notably recursive classes), it would likely add excess complexity, requiring subtyping (on N dimensions); untyped closures which can operate on types; and likely some other painpoints. And it's not clear how they could make the system unsafe as apart from being used to define recursive types are simply specialized versions of preexisting types.

\subsubsection{Functions}
Really the only reason I opted for the function overloading system instead of normal branching in the original language was in order to make it easier to extend the language.
Function overloading seems like it would be needlessly painful to formalize and doesn't contribute to our proof of soundness any more than branching via a ternary. As the case for functions having potentially no possible branch results in a type error in the complete language and this is simplified by having an else clause.

\section{Syntax}
Taking the ideas from the above rant into account we arrive at something along these lines
\[
\begin{array}{r l}
    \tau \bnfdef & \kwInt \bnfalt \kwLong \bnfalt \kwFloat \bnfalt \kwDouble \bnfalt 
        (\tau^*) \bnfalt \kwUnit \bnfalt \kwVoid \bnfalt \tau_1~ \tau_2~ + 
        \bnfalt ~ \kwArrow{\tau_1} {\tau_2} \bnfalt \kwType{e}\\

    \tau^* \bnfdef & \varepsilon \bnfalt \tau^* ~ \tau \bnfalt \kwunpack{\tau}\\

    e \bnfdef & e~e~+ \bnfalt ~e~e~\times \bnfalt ~ \kwtern{e}{e}{e} 
        \bnfalt e~e~== \bnfalt ~ \tau~ \tau~ ==
        \bnfalt ~ ( e^* )
        \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
        \bnfalt \kwid{x} \bnfalt \kwunesc{x}\\

    e^* \bnfdef & \varepsilon \bnfalt e^* ~ e \bnfalt \kwunpack{e}\\

    S \bnfdef & S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e

\end{array}
\]

Notes
\begin{itemize}
    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$
    \item notice that unpack can have any non-negative number of output values depending on the given input
    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
\end{itemize}

@@

\subsection{Findings, Resolution?, Conclusions}
It's entirely possible that we could encounter problems with the language design, although pretty unlikely since we're even requiring type signatures on closures.
